---
title: "Atl Herring Recruitment BRT"
author: "AM"
date: "2024-02-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Boosted Regression Trees for herring recruitment

Code for getting data, generating time series of possible predictors, plotting recruitment and predictor time series, producing correlation matrices to reduce the possible predictor list, and running boosted regression trees.

```{r packages, echo = F}
library(dplyr)
library(ggplot2)
library(corrplot)
library(reshape2)
library(ggstatsplot)
library(gbm)
library(tidyr)
library(dismo)
library(gridExtra)
library(ggcorrplot)
library(ggpubr)
```

## Load data

The data was modified, summarized, compiled and exported in a separate script

-   OISST:
-   Bottom Temperature: high resolution daily composite product from four ocean models (as in du Pontavice et al. 2023)
-   Environmental indices
    1.  Heatwave index
    2.  Cold Pool index
    3.  Gulf Stream Index
    4.  Warm Slope Water proportion
-   Food
    -   Abundance of calanus finmarchicus stage C5 in the GoM in fall
    -   Zooplankton abundance
        -   Cnidaria
        -   small calanoid
        -   large calanoid
-   Predators
    -   Haddock index: retrospective adjusted SSB from the GoM Haddock 2019 assessment (what about GB)
    -   Mackerel SSB is from the 2021 assessment
-   Biological indices
    -   Chlorophyll a
    -   Primary production
-   Herring recruitment indices from 2022 assessment

```{r data}
# Load in the combined data (mostly for plotting)
getwd()
dat <- read.csv(here::here('Data/Combined.Data.csv'))

# Load in the variable data and add lags
```

## Time Series

```{r recruitment time series}
dat%>% 
  dplyr::select(year,recruits,R.no.devs,logR.dev, SR.std.resid, SSB, SPR,Rs)%>%
  reshape2::melt(id.vars="year", variable.name="series")%>%
  ggplot(aes(year,value)) +
  geom_line()+ 
  theme_bw()+
  facet_wrap(~ series, scales = "free")
```

```{r environmental ts plots, echo = FALSE}

dat%>% 
  dplyr::select(year, WSW, CP, HW, BT, OISST) %>%
  reshape2::melt(id.vars="year", variable.name="series")%>%
  ggplot(aes(year,value)) +
  geom_line()+ 
  theme_bw()+
  facet_wrap(~ series, scales = "free", ncol=4)
```

```{r biological ts plots, echo = FALSE}
dat%>% 
  dplyr::select(year, HadSSB_GOM, HadSSB_GB, MackSSB, Jelly.Abun, Sm.cal.abun, Lg.cal.abun, Cfin.C5)%>%
  reshape2::melt(id.vars="year", variable.name="series")%>%
  ggplot(aes(year,value)) +
  geom_line(size=1)+ 
  theme_bw()+
  facet_wrap(~ series, scales = "free", ncol=4)
```

```{r}

```

```{r add lags }



```
